No one-size-fits-all solution to clean GBIF

### 2020-03-02 manuscript_peerj.Rmd
co-author version

    address: sDiv, German Center for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Leipzig, Germany

@@

@@

    address: Universidade Estadual de Feira de Santana, Bahia, Brazil

@@

  Species occurrence records provide the basis for many biodiversity studies. They derive from geo-referenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications. Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern. Automatic flagging and filtering are a scalable and reproducible means to identify potentially problematic records in datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org). However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses. Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 16 Neotropical taxa including plants and animals, terrestrial and marine, downloaded from GBIF. We find that 29-67% of the records are potentially erroneous, with large variation across taxonomic groups. Tests for duplicated information, collection year, basis of record as well as urban areas and political centroids have the greatest effect. While many flagged records might not be *de facto* erroneous, they could be overly imprecise and increase uncertainty in downstream analyses. Automated flagging can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus. Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research. 

@@

    @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o;@GBIForg2019p;@GBIForg2019q

@@

Manual data cleaning based on expert knowledge can detect these issues, but it is only feasible on small taxonomic or geographic scales, and it is time-consuming and difficult to reproduce. As an alternative, automated flagging methods to identify potentially problematic records have been proposed as a scalable option, as they are able to deal with datasets containing up to millions of records and many different taxa. Those methods are usually based on geographic gazetteers [e.g., @Chamberlain2016; @Zizka2019; @Jin2020] or on additional data, such as environmental variables [@Robertson2016]. Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biodiveristy research [@Zizka2019].

Erroneous records are especially problematic in conservation, where stakes are high. Recently proposed methods for automated conservation assessments could support the formal assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) [@Dauby2017; @Bachman2011; @Pelletier2018]. These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B ("Geographic range"). These methods have been used topropose preliminary global [@Stevart2019; @Zizka2019a] and regional [@Schmidt2017; @Cosiaux2018] Red List assessments. However, all metrics but especially EOO, are sensitive to individual records with erroneous coordinates **(REFERENCE)**. Automated conservation assessments may therefore be biased, particularly if the number of records are few, as it is the case for many tropical species.

While newly proposed methods for automated cleaning of species records hold great promise for biodiversity research, their use across taxonomic groups and datasets remains poorly explored. Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different groups of animals, funghi, and plants. Furthermore, we test the impact of automated filtering procedures for the accuracy of automated conservation assessments compared to full IUCN assessments. Specifically, we evaluate a pipeline of 13 automated filters to flag possibly erroneous records by using 1) record meta-data and 2) geographic gazetteers. We address three questions:

@@

# Material and Methods {-}
## Choice of study taxa {-}
This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering graduate students, taxonomists and biogeographers working with different taxonomic groups of animals, funghi, and plants across the Neotropics. Each participant analysed geographic occurrence data from their taxonomic group of interest and provided an informed assessment of the results for their group. Hence, we include groups based on the expertise of the participants wrather than following some arbitrary choice of taxa and taxonomic ranks. We acknowledge a varying degree in documented expertise and number of years working on each group. We obtained public occurrence records for 16 taxa, including one plant family, eight plant genera, one genus of fungi, two families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).

@@

We downloaded occurrence information for all study groups from www.gbif.org using the `rgbif` v1.4.0 package [@Chamberlain2017] in R (GBIF.org, 2019a-p). We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33$^\circ$ N and longitudes smaller than 35$^\circ$ W and larger than 120$^\circ$ W reflecting the Neotropics, our main area of expertise [@Morrone2014]. The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, *Gaylussacia*, Iridaceae, Neanuridae, and *Tillandsia* for which we only obtained the Neotropical occurrences.

##  Automated cleaning {-}
We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99. Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e. test for duplicated records). In a second step, we used the `clean_coordinates` function of the `CoordinateCleaner v2.0-11` package [@Zizka2019] with default options to flag errors that are common to biological data sets (“filtered”, hereafter). These including: coordinates in the sea for terrestrial taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions. See Table 2 for a summary of all flagging tests we used.

@@

## Evaluation of results {-}
<!-- This session is exciting and makes the study more reliable and interesting than one just based on data mining, but how this was done is very unclear. Did people provide a specific score for those, or how were their assessments recorded? I think we need to clarify this, as it’s the weakest part of the paper at the moment. We don’t report the results for this evaluation nor what they mean.  -->
Each specialist provided an informed assessment on the performance of the raw and cleaned datasets, respectively concerning: 1) the number of flagged records; 2) the importance of specific tests; 3) the accuracy of the overall species richness maps; 4) the accuracy of the preliminary conservation assessments.

We did all analyses in the R computing environment [@rcoreteam2019], using standard libraries for data handling and visualization [@Wickham2018; @Garnier2018; @Ooms2014; @Ooms2019; @Hijmans2019]. All scripts are available from https://github.com/idiv-biodiversity/effects_of_automated_cleaning, or from the Appendix 1 in the electronic supplement of this study.

@@

# Table 2
tab2 <- read_csv("output/summary_table_flags.csv") %>% 
  dplyr::select(-coordinate_missing, -invalid_coords, -gbif_coords,-inst_coords, -equal_coords) %>% 
  mutate(fraction = round(fraction * 100,1))

tab2 <- tab2[,c(1:3,14,5,7,4,8,6,10,13,11,12,9)]

@@

tab2 <- tab1 %>% 
  select(taxon) %>%  
  left_join(tab2, by = "taxon") %>% 
  replace_na(list(total_records = 0))

@@

We retrieved a total of `r nrow(dat)` species occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3, Appendix 2). We obtained most records for `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(taxon) %>%  unlist()` (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs) %>%  unlist()`) and fewest for *`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(taxon) %>%  unlist()`* (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(nrecs) %>%  unlist()`). The species with most records was *`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(species)`* (`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs)`).

Our automated tests flagged a total of `r sum(!dat$summary)` records (Fig. \ref{fig:total}), with a median of `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per group. Overall, the most important test was for duplicated records (on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!.dpl)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per taxonomic group). The filtering steps based on record meta-data that flagged the largest number of records were the basis of records (`r dat %>% group_by(taxon) %>% summarize(num = mean(!coordinate_base)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the collection year (`r dat %>% group_by(taxon) %>% summarize(num = mean(!record_age)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%). The most important automated tests were for urban area (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.urb)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%), the capital coordinates (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.cap)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the centroids of political units (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.cen)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%, see Table 3 and Supplementary material 1 for further details and the absolute numbers).  Only few records were flagged by the zero coordinates and biodiversity institution tests (Fig. \ref{fig:split}), and no records were flagged by the sea test.

*`r tab2 %>% arrange(fraction) %>% top_n(1) %>% select(taxon) %>% unlist()`*, *`r tab2 %>% arrange(fraction) %>% top_n(3) %>% select(taxon) %>% slice(2) %>% unlist()`*, and *`r tab2 %>% arrange(fraction) %>% top_n(3) %>% select(taxon) %>% slice(3) %>% unlist()`* had the highest fraction of flagged records (Table 3). In general, the different filters we tested were of similar importance for different study groups. There were few outstanding exceptions, including the particularly high proportions of records flagged on the "basis of record test" for *Tityus* (7.0%), Dipsadidae (5.6%), *Prosthechea* (5.0%) and *Tillandsia* (4.9%), by the collection year for Dipsadidae (11.3%), by the taxonomic identification level for  *Tityus* (1.6%), by the capital coordinates for *Oocephalus* (6.1%) and *Gaylussacia* (3.2%), and by the urban areas test for *Oocephalus* (13.3%) and Iridaceae (12.3%).

Geographically, the records flagged by the "basis of record" and "individual count" tests were concentrated in Central America and southern North America, and a relatively high number of records flagged due to their proximity to the centroids of political entities were located on Caribbean islands (Fig. \ref{fig:split}). See Supplementary material 2 for species richness maps using the raw and cleaned data for all study groups.

We found IUCN assessments for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>% nrow()` species that were also included in our distribution data from 10 of our study groups (Table 4, Appendix 3). The fraction of species evaluated varied among the study group, with a maximum of `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  max()`% for *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>% slice(1L) %>% select(taxon) %>%  unlist()`* and *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>%  select(taxon) %>% slice(2) %>% unlist()`* and a minimum of `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(iucn_frac_eval) %>%  unlist()`% for `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(taxon) %>%  unlist()` (note that the number of species varied considerably among groups). The median percentage of species per study group with an IUCN assessment was `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  median()`%. A total of `r aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "possibly_threatened") %>% select(test) %>% unlist() %>%  sum()` species were listed as Threatened by the IUCN global Red List (CR = `r aa %>% filter(iucn_category == "CR") %>% nrow()`, EN = `r aa %>% filter(iucn_category == "EN") %>% nrow()`, VU = `r aa %>% filter(iucn_category == "VU") %>% nrow()`) and `r  aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "not_threatened") %>% select(test) %>% unlist() %>%  sum()` as Not Threatened.

We obtained automated conservation assessments for `r aa %>% select(flag_Category_CriteriaB) %>% filter(!is.na(flag_Category_CriteriaB)) %>% nrow()` species in the filtered dataset. Based on the filtered data, the automated conservation assessment evaluated `r sum(aa$flag_status == "possibly_threatened", na.rm = TRUE)` species as possibly threatened (`r round(sum(aa$flag_status == "possibly_threatened", na.rm=TRUE) / aa %>% filter(!is.na(flag_status)) %>% nrow() * 100, 1)`%, CR = `r aa %>% filter(flag_Category_CriteriaB == "CR") %>% nrow()`, EN = `r aa %>% filter(flag_Category_CriteriaB == "EN") %>% nrow()`, VU = `r aa %>% filter(flag_Category_CriteriaB == "VU") %>% nrow()`). The automated assessment based on the filtered dataset agreed with the IUCN assessment for identifying species as possibly threatened (CR, EN, VU) for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status)) %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status))  %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%; Table 4). Filtering reduced the EOO by `r tab3 %>%  select(flag_median_EOO_change) %>%  unlist() %>%  median()`% and the AOO by `r tab3 %>%  select(flag_median_AOO_change) %>%  unlist() %>%  median()`% on median per group. For the raw dataset the agreement with IUCN was higher at `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%).

# Discussion {-}
<!-- add some subheadings to discussion -->
Automated flagging based on meta-data and automatic tests flagged on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group. The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (Question 1). The importance of different tests was similar across taxonomic groups, with few exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (Question 2). The results for species richness were similar between the raw and filtered data with some improvements by the filters. We found little impact of filtering on the accuracy of the automated conservation assessments (Question 3).

## The relevance of individual filters {-}
Several reasons may explain the high proportion of records flagged as duplicates. First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a specimen duplication is easily feasible. Second, independent collections at similar localities may occur, in particular for local endemics. Third, low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities. Duplicated records do not represent errors per se, but an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.

>>>>

    address: sDiv, German Center for Integrative Biodiversity Research Halle-Jena-Leipzig (iDiv), Leipzig, Germany

@@

    address: Universidade Estadual de Feira de Santana, Feira de Santana, Brazil

@@

@@

  Species occurrence records provide the basis for many biodiversity studies. They derive from geo-referenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications. Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern. Automatic flagging and filtering are a scalable and reproducible means to identify potentially problematic records in datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org). However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses. Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 18 Neotropical taxa including animals, fungi, and plants, terrestrial and marine, downloaded from GBIF. We find that 29-90% of the records are potentially erroneous, with large variation across taxonomic groups. Tests for duplicated information, collection year, basis of record as well as urban areas and coordinates for terrestrial taxa in the sea or marine taxa on land have the greatest effect. While many flagged records might not be *de facto* erroneous, they could be overly imprecise and increase uncertainty in downstream analyses. Automated flagging can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus. Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research. 

@@

    @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o;@GBIForg2019p;@GBIForg2020, @GBIForg2020a

@@

Manual data cleaning based on expert knowledge can detect these issues, but it is only feasible on small taxonomic or geographic scales, and it is time-consuming and difficult to reproduce. As an alternative, automated flagging methods to identify potentially problematic records have been proposed as a scalable option, as they are able to deal with datasets containing up to millions of records and many different taxa. Those methods are usually based on geographic gazetteers [e.g., @Chamberlain2016; @Zizka2019; @Jin2020] or on additional data, such as environmental variables [@Robertson2016]. Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the suitability of publicly available occurrence records for biodiversity research [@Zizka2019].

Erroneous records are especially problematic in conservation, where stakes are high. Recently proposed methods for automated conservation assessments could support the formal assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) [@Dauby2017; @Bachman2011; @Pelletier2018]. These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B ("Geographic range"). These methods have been used to propose preliminary global [@Stevart2019; @Zizka2019a] and regional [@Schmidt2017; @Cosiaux2018] Red List assessments. However, all metrics but especially EOO, are sensitive to individual records with erroneous coordinates. Automated conservation assessments may therefore be biased, particularly if the number of records are few, as it is the case for many tropical species.

While newly proposed methods for automated cleaning of species records hold great promise for biodiversity research, their use across taxonomic groups and datasets remains poorly explored. Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different groups of animals, fungi, and plants. Furthermore, we test the impact of automated filtering procedures for the accuracy of automated conservation assessments compared to full IUCN assessments. Specifically, we evaluate a pipeline of 13 automated filters to flag possibly erroneous records by using 1) record meta-data and 2) geographic gazetteers. We address three questions:

@@

# Material and Methods {-}
## Choice of study taxa {-}
This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering students and researchers working with different taxonomic groups of animals, fungi, and plants across the Neotropics. Each participant analysed geographic occurrence data from their taxonomic group of interest and commented on the results for their group. Hence, we include groups based on the expertise of the participants rather than following some arbitrary choice of taxa and taxonomic ranks. We acknowledge a varying degree in documented expertise and number of years working on each group. We obtained public occurrence records for 18 taxa, including one plant family, nine plant genera, one genus of fungi, three families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).

@@

We downloaded occurrence information for all study groups from www.gbif.org using the `rgbif` v1.4.0 package [@Chamberlain2017] in R (GBIF.org, 2019a-p,2020a,b). We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33$^\circ$ N and longitudes smaller than 35$^\circ$ W and larger than 120$^\circ$ W reflecting the Neotropics, our main area of expertise [@Morrone2014]. The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, *Gaylussacia*, Iridaceae, Neanuridae, and *Tillandsia* for which we only obtained the Neotropical occurrences.

##  Automated cleaning {-}
We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99. Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e., test for duplicated records). In a second step, we used the `clean_coordinates` function of the `CoordinateCleaner v2.0-11` package [@Zizka2019] with default options to flag errors that are common to biological data sets (“filtered”, hereafter). These include: coordinates in the sea for terrestrial taxa and on land for marine taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions. See Table 2 for a summary of all flagging tests we used.

@@

## Evaluation of results {-}
<!-- This session is exciting and makes the study more reliable and interesting than one just based on data mining, but how this was done is very unclear. Did people provide a specific score for those, or how were their assessments recorded? I think we need to clarify this, as it’s the weakest part of the paper at the moment. We don’t report the results for this evaluation nor what they mean.  -->
Each author provided an informed comment on the performance of the raw and cleaned datasets, respectively concerning the number of flagged records and the accuracy of the overall species richness maps and we compared the agreement between automated conservation assessments based on raw and filtered occurrences with the global IUCN Red List (www.iucn.org).

We did all analyses in the R computing environment [@rcoreteam2019], using standard libraries for data handling and visualization [@Wickham2018; @Garnier2018; @Ooms2014; @Ooms2019; @Hijmans2019]. All scripts are available from a zenodo repository (**DOI**).

@@

# Table 2
tab2 <- read_csv("output/summary_table_flags.csv") %>% 
  dplyr::select(-coordinate_missing, -invalid_coords, -gbif_coords, -equal_coords) %>% 
  mutate(fraction = round(fraction * 100,1))

tab2 <- tab2[,c(1:3,16,5,7,4,8,6,14,10,15,11:13,9)]

@@

tab2 <- tab1 %>% 
  select(taxon) %>%  
  bind_rows(data.frame(taxon = "Total")) %>% 
  left_join(tab2, by = "taxon") %>% 
  replace_na(list(total_records = 0))

@@

We retrieved a total of `r nrow(dat)` species occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3, Appendix 1). We obtained most records for `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(taxon) %>%  unlist()` (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs) %>%  unlist()`) and fewest for *`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(taxon) %>%  unlist()`* (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(nrecs) %>%  unlist()`). The species with most records was *`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(species)`* (`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs)`).

Our automated tests flagged a total of `r sum(!dat$summary)` records (Fig. \ref{fig:total}), with a median of `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per group. Overall, the most important test was for duplicated records (on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!.dpl)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per taxonomic group). The filtering steps based on record meta-data that flagged the largest number of records were the basis of records (`r dat %>% group_by(taxon) %>% summarize(num = mean(!coordinate_base)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the collection year (`r dat %>% group_by(taxon) %>% summarize(num = mean(!record_age)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%). The most important automated tests were for urban area (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.urb)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the occurrence from records of terrestrial taxa in the sea and marine taxa on land (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.sea)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%, see Table 3 and Appendix 1 in the electronic supplement for further details and the absolute numbers).  Only few records were flagged by the coordinate precision, zero coordinates and biodiversity institution tests (Fig. \ref{fig:split}).

*`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(1) %>% unlist()`*, *`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(2) %>% unlist()`*, and *`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(3) %>% unlist()`* had the highest fraction of flagged records (Table 3). In general, the different filters we tested were of similar importance for different study groups. There were few outstanding exceptions, including the particularly high proportions of records flagged on the "basis of record test" for *Tityus* (7.0%), Dipsadidae (5.6%), *Prosthechea* (5.0%) and *Tillandsia* (4.9%), by the collection year for Dipsadidae (11.3%), by the taxonomic identification level for  *Tityus* (1.6%), by the capital coordinates for *Oocephalus* (6.1%) and *Gaylussacia* (3.2%), by the seas/land test for Diogenidae and *Thozetella*, and by the urban areas test for *Oocephalus* (13.3%) and Iridaceae (12.3%). Furthermore, Entomobryidae differed considerably from all other study taxa with exceptionally high numbers of records flagged by the "basis of record, "level of identification' and urban areas" tests.

Geographically, the records flagged by the "basis of record" and "individual count" tests were concentrated in Central America and southern North America, and a relatively high number of records flagged due to their proximity to the centroids of political entities were located on Caribbean islands (Fig. \ref{fig:split}). See Appendix 2 for species richness maps using the raw and cleaned data for all study groups.

We found IUCN assessments for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>% nrow()` species that were also included in our distribution data from 11 of our study groups (Table 4, Appendix 3). The fraction of species evaluated varied among the study group, with a maximum of `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  max()`% for *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>% slice(1L) %>% select(taxon) %>%  unlist()`* and *`r tab3 %>% arrange(desc(iucn_frac_eval)) %>%  select(taxon) %>% slice(2) %>% unlist()`* and a minimum of `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(iucn_frac_eval) %>%  unlist()`% for `r tab3 %>% arrange(iucn_frac_eval) %>% slice(1) %>%  select(taxon) %>%  unlist()` (note that the number of total species varied considerably among groups). The median percentage of species per study group with an IUCN assessment was `r tab3 %>% select(iucn_frac_eval) %>% unlist() %>%  median()`%. A total of `r aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "possibly_threatened") %>% select(test) %>% unlist() %>%  sum()` species were listed as Threatened by the IUCN global Red List (CR = `r aa %>% filter(iucn_category == "CR") %>% nrow()`, EN = `r aa %>% filter(iucn_category == "EN") %>% nrow()`, VU = `r aa %>% filter(iucn_category == "VU") %>% nrow()`) and `r  aa %>% filter(!is.na(iucn_category)) %>%  filter(iucn_category != "DD") %>%filter(iucn_category != "EX") %>%  mutate(test = iucn_status == "not_threatened") %>% select(test) %>% unlist() %>%  sum()` as Not Threatened.

We obtained automated conservation assessments for `r aa %>% select(flag_Category_CriteriaB) %>% filter(!is.na(flag_Category_CriteriaB)) %>% nrow()` species in the filtered dataset. Based on the filtered data, the automated conservation assessment evaluated `r sum(aa$flag_status == "possibly_threatened", na.rm = TRUE)` species as possibly threatened (`r round(sum(aa$flag_status == "possibly_threatened", na.rm=TRUE) / aa %>% filter(!is.na(flag_status)) %>% nrow() * 100, 1)`%, CR = `r aa %>% filter(flag_Category_CriteriaB == "CR") %>% nrow()`, EN = `r aa %>% filter(flag_Category_CriteriaB == "EN") %>% nrow()`, VU = `r aa %>% filter(flag_Category_CriteriaB == "VU") %>% nrow()`, see Appendix 3 for assessments of all species). The automated assessment based on the filtered dataset agreed with the IUCN assessment for identifying species as possibly threatened (CR, EN, VU) for `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status)) %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% filter(iucn_category != "EX") %>%  filter(!is.na(flag_status))  %>% mutate(agreement = iucn_status == flag_status) %>% select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%; Table 4). Filtering reduced the EOO by `r tab3 %>%  select(flag_median_EOO_change) %>%  unlist() %>%  median()`% and the AOO by `r tab3 %>%  select(flag_median_AOO_change) %>%  unlist() %>%  median()`% on median per group. For the raw dataset the agreement with IUCN was higher at `r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% select(agreement) %>% unlist() %>% sum() ` species (`r aa %>% filter(!is.na(iucn_category)) %>% filter(iucn_category != "DD") %>% mutate(agreement = iucn_status == raw_status) %>% select(agreement) %>% unlist() %>% mean() %>%  round(3) * 100`%).

# Discussion {-}
<!-- add some subheadings to discussion -->
Automated flagging based on meta-data and automatic tests flagged on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group. The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (Question 1). The importance of different tests was similar across taxonomic groups, with exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (Question 2). The results for species richness were similar between the raw and filtered data with some improvements by the filters. We found little impact of filtering on the accuracy of the automated conservation assessments (Question 3).

## The relevance of individual filters {-}
Several reasons may explain the high proportion of records flagged as duplicates. First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a specimen duplication is easily feasible. Second, independent collections at similar localities may occur, in particular for local endemics. Third, low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities. Duplicated records do not represent errors *per se*, but an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.
